Supplementary Content for the article *Functional diversity can facilitate the collapse of an undesirable ecosystem state*

### 2022-06-08 experiments/2_supplement/supplement.Rmd
changes supplement

---
title: Supplementary Content for the article *The role of functional diversity in ecosystem regime shifts*
author: "Owen L Petchey, Uriah Daugaard, Anubhav Gupta, Rainer M. Krug, Kimberley D. Lemmen, Sofia J. van Moorsel, Marcel Suleiman, Debra Zuppinger-Dingley, Romana Limberger"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
---

```{r include = FALSE, echo = FALSE}
knitr::opts_chunk$set(include = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      cache = FALSE)
```

@@

Here we provide additional description of the method used to create strain variation within functional groups. The following should be read only after the relevant text in the main report.  We generated variation in maximum growth rate $G_{max}$ and tolerance to reduced sulfur $h_{SR}$ or oxygen $h_{O}$ by varying the range of trait values around a reference trait value. Take for example the cyanobacteria functional group, a reference maximum growth rate parameter $G_{max,CB}$ and a reference tolerance parameter $h_{SR,CB}$. Recall that amount of among strain variation is controlled by the two meta-parameters $Div_{G_{max,CB}}$ and $Div_{h_{SR,CB}}$. To create a trade-off, the two meta-parameters were always of different sign. The among strain variation was calculated such that the growth rate of strain $i = 1$ was a factor $1/({2Div_{G_{max,CB}}})$ of the reference growth rate, and the growth rate of the $i = n$ strain was a factor $2Div_{G_{max,CB}}$ of the reference growth rate. The growth rate of the $i = {2,...n-1}$ other strains was the reference growth rate multiplied by a factor that was equally distributed between the factor of the $i = 1$ and $i = n$ strain. This implementation of the variation and trade-off has the assumption of a linear trade-off of log-log transformed among-strain variation. Hence, for example, with the reference growth rate $G_{max,CB} = 0.05$, among strain variation of $Div_{G_{max,CB}} = 1$, and nine strains ($n = 9$), the growth rates of the nine strains are 0.025, 0.03, 0.035, 0.042, 0.05, 0.059, 0.071, 0.084, 0.1. Put another way, the growth rate parameter of the $i$-th of the $n$ strains in a functional group was calculated as $G_{max,i} = D_i * G_{max}$ where $D_i$ is the $i$-th element of the set $D = {2f(x)Div_{G_{max}} |x ∈ N, x ≤ n}$, where $f(x) = (x − (n + 1)/2)/((n − 1)/2)$.

@@

(ref:div-repl-meth1) **Replication method results.** **Left column of panels**: the effect of response trait diversity (y-axis) on the positions of the tipping points (anoxic to oxic in blue, oxic to anoxic in red), which determine the extent of the region of bistability (green shaded region) in the oxygen diffusivity gradient. The left-most region is anoxic only, the middle is oxic or anoxic depending on historical conditions, and the right-most is oxic only. The thick grey vertical lines at -8 and 0 on the x-axis show the extent of the oxygen diffusivity gradient investigated in the simulations. **Right column of panels**: The effect of diversity on resilience, with positive values on the y-axis indicating that diversity has increased resilience, i.e. delayed the shift to the other state.

@@

(ref:div-temp-eventdef1) **Results with an abundance floor.** **Left column of panels**: the effect of response trait diversity (y-axis) on the positions of the tipping points (anoxic to oxic in blue, oxic to anoxic in red), which determine the extent of the region of bistability (green shaded region) in the oxygen diffusivity gradient. The left-most region is anoxic only, the middle is oxic or anoxic depending on historical conditions, and the right-most is oxic only. The thick grey vertical lines at -8 and 0 on the x-axis show the extent of the oxygen diffusivity gradient investigated in the simulations. **Right column of panels**: The effect of diversity on resilience, with positive values on the y-axis indicating that diversity has increased resilience, i.e. delayed the shift to the other state.

@@

The results below show the stable states as a function of oxygen diffusivity for a system with an intermediate level of diversity in all functional groups, with either 1'000'000 hours at each oxygen diffusivity level (Figure \@ref(fig:ss-temp-meth3)), or 2'000'000 hours (Figure \@ref(fig:ss-temp-meth4)). This shows that 1'000'000 hours duration of the simulations are sufficient to reach stable state, except very close to state transitions on the trajectory of recovery (i.e.: cyanobacteria: trajectory up; sulfate-reducing bacteria: trajectory down). Very close to these transitions, even at 2'000'000 hours, the dynamics are still transient, with high tolerance strains at high density, even though they will eventually be outcompeted by low tolerance (higher growth rate) strains.

@@

When the environment ameliorates for the sulfate-reducing bacteria (i.e. trajectory down), there is only limited evidence of strain replacement (in final state) along the oxygen diffusivity gradient -- either the system is oxic or the least tolerant strain SB9 dominates. And yet the switch to anoxic state occurs earlier than when there is no diversity, which suggests there is some role of the more tolerant strains. We here show that the more tolerant strains do play a role, but only a transient one. The following dynamics are for the system starting oxic, and with a value of oxygen diffusivity (log10(oxygen diffusivity) = -4.8) for which the system remains oxic if there is no diversity (Figure \@ref(fig:sb-div-effect3)), but switches to anoxic if there is diversity (Figure \@ref(fig:sb-div-effect4)). There is only diversity in the two groups of sulfur bacteria. The most tolerant strain does at first grow, but is then outcompeted by less tolerant strains as the concentration of the inhibiting substrate (oxygen) declines.

@@

@@

>>>>

---
title: Supplementary Content for the article *Functional diversity can facilitate the collapse of an undesirable ecosystem state*
author: "Owen L Petchey, Uriah Daugaard, Anubhav Gupta, Rainer M. Krug, Kimberley D. Lemmen, Sofia J. van Moorsel, Marcel Suleiman, Debra Zuppinger-Dingley, Romana Limberger"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
---

```{r include = FALSE, echo = FALSE}
knitr::opts_chunk$set(include = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      cache = TRUE)
```

@@

Here we provide additional description of the method used to create strain variation within functional groups. The following should be read only after the relevant text in the main report.  We generated variation in maximum growth rate $G_{max}$ and tolerance to reduced sulfur $h_{SR}$ or oxygen $h_{O}$ by varying the range of trait values around a reference trait value. To manipulate the amount of variation in a given trait and functional group, we introduced the meta-parameter $Div$ subscripted by the trait and the functional group. Take for example the cyanobacteria functional group, a reference maximum growth rate parameter $G_{max,CB}$ and a reference tolerance parameter $h_{SR,CB}$. The amount of among strain variation in the cyanobacteria is controlled by the two meta-parameters $Div_{G_{max,CB}}$ and $Div_{h_{SR,CB}}$. To create a trade-off, the two meta-parameters were always of different sign. The among strain variation was calculated such that the growth rate of strain $i = 1$ was a factor $1/({2Div_{G_{max,CB}}})$ of the reference growth rate, and the growth rate of the $i = n$ strain was a factor $2Div_{G_{max,CB}}$ of the reference growth rate. The growth rate of the $i = {2,...n-1}$ other strains was the reference growth rate multiplied by a factor that was equally distributed between the factor of the $i = 1$ and $i = n$ strain. This implementation of the variation and trade-off has the assumption of a linear trade-off of log-log transformed among-strain variation. Hence, for example, with the reference growth rate $G_{max,CB} = 0.05$, among strain variation of $Div_{G_{max,CB}} = 1$, and nine strains ($n = 9$), the growth rates of the nine strains are 0.025, 0.03, 0.035, 0.042, 0.05, 0.059, 0.071, 0.084, 0.1. Put another way, the growth rate parameter of the $i$-th of the $n$ strains in a functional group was calculated as $G_{max,i} = D_i * G_{max}$ where $D_i$ is the $i$-th element of the set $D = {2f(x)Div_{G_{max}} |x ∈ N, x ≤ n}$, where $f(x) = (x − (n + 1)/2)/((n − 1)/2)$.

@@

(ref:div-repl-meth1) **Replication method results.** **Left column of panels**: the effect of trait diversity (y-axis) on the positions of the tipping points (anoxic to oxic in blue, oxic to anoxic in red), which determine the extent of the region of bistability (green shaded region) in the oxygen diffusivity gradient. The left-most region is anoxic only, the middle is oxic or anoxic depending on historical conditions, and the right-most is oxic only. The thick grey vertical lines at -8 and 0 on the x-axis show the extent of the oxygen diffusivity gradient investigated in the simulations. **Right column of panels**: The effect size of trait variation on resilience, with positive values on the y-axis indicating that diversity has increased resilience.

@@

(ref:div-temp-eventdef1) **Results with an abundance floor.** **Left column of panels**: the effect of trait diversity (y-axis) on the positions of the tipping points (anoxic to oxic in blue, oxic to anoxic in red), which determine the extent of the region of bistability (green shaded region) in the oxygen diffusivity gradient. The left-most region is anoxic only, the middle is oxic or anoxic depending on historical conditions, and the right-most is oxic only. The thick grey vertical lines at -8 and 0 on the x-axis show the extent of the oxygen diffusivity gradient investigated in the simulations. **Right column of panels**: The effect size of trait variation on resilience, with positive values on the y-axis indicating that diversity has increased resilience, i.e. delayed the shift to the other state.

@@

The results below show the stable states as a function of oxygen diffusivity for a system with an intermediate level of diversity in all functional groups, with either 1'000'000 hours at each oxygen diffusivity level (Figure \@ref(fig:ss-temp-meth3)), or 2'000'000 hours (Figure \@ref(fig:ss-temp-meth4)). This shows that 1'000'000 hours duration of the simulations are sufficient to reach stable state, except very close to state transitions on the trajectory of recovery (i.e.: cyanobacteria: trajectory of increasing oxygen diffusivity; sulfate-reducing bacteria: trajectory of decreasing oxygen diffusivit). Very close to these transitions, even at 2'000'000 hours, the dynamics are still transient, with high tolerance strains at high density, even though they will eventually be outcompeted by low tolerance (higher growth rate) strains.

@@

When the environment ameliorates for the sulfate-reducing bacteria (i.e. trajectory of decreasing oxygen diffusivity), there is only limited evidence of strain replacement (in final state) along the oxygen diffusivity gradient -- either the system is oxic or the least tolerant strain dominates. And yet the switch to anoxic state occurs at higher oxygen diffusivity than when there is no diversity, which suggests there is some role of the more tolerant strains. We here show that the more tolerant strains do play a role, but only a transient one. The following dynamics are for the system starting oxic, and with a value of oxygen diffusivity (log10(oxygen diffusivity) = -4.8) for which the system remains oxic if there is no diversity (Figure \@ref(fig:sb-div-effect3)), but switches to anoxic if there is diversity (Figure \@ref(fig:sb-div-effect4)). There is only diversity in the two groups of sulfur bacteria. The most tolerant strain does at first grow, but is then outcompeted by less tolerant strains as the concentration of the inhibiting substrate (oxygen) declines.

@@

Among these two pathways, effects of phototrophic sulfur bacteria via abiotic oxidation of sulfide are probably of subordinate importance in natural environments because abiotic oxidation rates of sulfide are very small compared to biotic oxidation rates ([Luther et al 2011]( https://doi.org/10.3389/fmicb.2011.00062)).

@@

# References
Bush, T., Diao, M., Allen, R.J., Sinnige, R., Muyzer, G. & Huisman, J. (2017). Oxic-anoxic regime shifts mediated by feedbacks between biogeochemical processes and microbial community dynamics. Nat. Commun., 8, 789.

Luther, G.W., Findlay, A., MacDonald, D., Owings, S., Hanson, T., Beinart, R., et al. (2011). Thermodynamics and kinetics of sulfide oxidation by oxygen: a look at inorganically controlled reactions and biologically mediated processes in the environment. Front. Microbiol., 2, 62.
